---
title: "Analyzing Melbourne’s Housing Market: Price Volatility and Growth Trends (2013-2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{property-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MelbournePrices)
library(tidyverse)
library(readr)
library(knitr)
library(readxl)
```

```{r}
load ("C:/Users/aalch/OneDrive/Desktop/Communicating with data/assignment-4-packages-and-shiny-apps-Saumya6915/data/property_data_wide.rda")

```


## <u> Introduction </u>

Melbourne's housing market has experienced significant shifts over the past decade. Once stable suburbs are now booming, and previously overlooked areas have become hotspots for investment. The dataset in this package allows us to dive deeply into these changes, revealing which suburbs have demonstrated robust growth, fluctuating volatility, and promising trends.

This vignette offers an analysis guide using data from 2013 to 2023, sourced from the [Victoria State Government website](https://www.land.vic.gov.au/valuations/resources-and-reports/property-sales-statistics). Through this analysis, we’ll identify suburbs with high investment potential and those with notable volatility, providing insights into the dynamics that drive Melbourne’s property market.

At some point, nearly everyone living in Melbourne has found themselves caught in the search for a place to buy, navigating the complexities of the city's housing market. But can you imagine what property prices looked like a decade ago? With Melbourne's real estate landscape evolving so dramatically, many are left wondering: which suburbs offer the best opportunities for investment, lifestyle, and long-term growth?

Dataset Description
The dataset provides several key metrics on property prices and growth in Melbourne. Here’s a breakdown of the variables:

## <u>Data Overview</u>

The data for this analysis was sourced from the [Victoria State Government website](https://www.land.vic.gov.au/valuations/resources-and-reports/property-sales-statistics) , covering the period from 2013 to 2023 @tbl-data-dictionary. 

The **Valuer-General Victoria (VGV)** releases comprehensive data which is compiled using information lodged every time a property sale is completed. It focuses on median house prices across various suburbs in Melbourne. The key metrics include annual median prices and percentage changes for each suburb.


```{r echo =FALSE}
#| label: tbl-data-dictionary

library(knitr)
library(kableExtra)

data_desc <- data.frame(
  Variable_Name = c("Locality", "Prelim", "22-23", "13-23", "PA", "year", "median_price"),
  Description = c(
    "The name of the suburb or locality in Melbourne.",
    "Preliminary house price estimate for the latest year.",
    "Percentage change in median house price from 2022 to 2023.",
    "Percentage growth in house prices from 2013 to 2023.",
    "Compound annual growth rate (CAGR) or price appreciation (PA) percentage.",
    "The year associated with the median price.",
    "The median house price in AUD for the specified locality and year."
  )
)

# Print the data dictionary table
kable(data_desc, caption = "Variables in the dataset") %>%
  kable_styling()

```

## <u>Variable Insights</u>

**Locality:** This variable helps us track property trends within specific suburbs, making it possible to identify high-growth areas.

**Prelim:** Preliminary data for the latest year gives early insights into price trends and can highlight shifts that may continue into future years.

**22-23 and 13-23:** These metrics provide percentage growth over one year (2022–2023) and over a decade (2013–2023), allowing us to understand both recent shifts and long-term trends.

**PA (CAGR):** Represents the annualized growth rate over ten years, smoothing out year-to-year volatility and reflecting the underlying trend.

**Year and median_price:** These columns form the basis of our time-series analysis, enabling us to examine price changes across years for each locality.

# <u>Understanding the Dynamics</u>

**1. Annual Price Growth for a Locality**

The annual_price_growth function is designed to help users track the year-over-year growth in median property prices for a specific suburb. This function is useful for examining trends within individual localities, helping users understand how property prices have evolved over time in response to factors such as demand, local development, and changes in infrastructure.

By calculating the year-over-year percentage change in median prices, this function reveals periods of growth, stagnation, or decline within a specific suburb. It is particularly valuable for investors, real estate professionals, and policymakers who want to assess the stability and growth potential of specific areas in Melbourne.

```{r}
annual_growth <- annual_price_growth(property_data, "ABBOTSFORD")
head(annual_growth)
```

The output displays the year, median_price, and annual_growth_rate columns for ABBOTSFORD. The annual_growth_rate column shows the percentage change in median property prices from one year to the next, giving insight into growth trends and potential volatility.

**2. Top-Performing Suburbs Based on Average Annual Growth **

This function identifies the top n localities with the highest price volatility, which can be useful for risk analysis. By calculating the average annual growth rate in property prices for each suburb over a specified period, this function provides insights into which suburbs have demonstrated strong and consistent price appreciation. Such information can guide investment decisions, highlight emerging neighborhoods, and shed light on property market trends.

```{r}
# Identify top 10 suburbs with the highest growth from 2013 to 2023

top_suburbs <- top_performing_suburbs(property_data, 2013, 2023, top_n = 10)
top_suburbs
```

This command will output a list of the top 10 suburbs ranked by their average annual growth rates in median house prices. Each suburb's growth rate represents the average year-over-year change in price, expressed as a percentage. For example, if a suburb shows an annual growth rate of 8%, it suggests that on average, property prices in that suburb have increased by 8% each year over the specified period.

**3. Calculating Price Volatility for a Specific Locality **

The price_volatility function assesses the variability of property prices within a specified locality. This function is valuable for understanding price stability, where high volatility may point to market speculation or changing demand, while low volatility often indicates a more stable market.

By calculating the standard deviation of median property prices over the available years, this function offers insights into price stability in a locality. Markets with higher volatility may present both greater risks and potential rewards, while stable markets are generally more attractive to conservative investors.

```{r}
volatility <- price_volatility(property_data, "ABBOTSFORD")
volatility
```

The output will display the Locality and Price_Volatility, where Price_Volatility represents the standard deviation of median property prices for ABBOTSFORD.

#### How to interpret   

*High Volatility:* A high Price_Volatility suggests significant yearly price fluctuations, often due to changing demand, speculative investments, or shifts in neighborhood appeal.

*Low Volatility:* Low volatility indicates stable prices, which can be attractive to long-term investors seeking consistent returns, often reflecting steady demand and established infrastructure.

For example, high volatility in ABBOTSFORD may signal substantial price swings due to speculative interest or new developments, while low volatility would imply more predictable returns.


**4. Summarizing Market Trends by Year **

The market_trends_by_year function provides a yearly summary of key property market metrics, such as the average median price, average growth rate, and the number of unique localities. This function helps users understand general trends over time, making it valuable for high-level analysis of property price patterns and market coverage in Melbourne.

```{r}
annual_trends <- market_trends_by_year(property_data)
head(annual_trends)
```

The output will show yearly data for avg_median_price, avg_growth_rate, and locality_count. This allows users to observe the market’s general growth, stability, and data coverage over the years.

#### How to interpret

*Average Median Price (avg_median_price):* A steady increase signals strong demand, while dips may indicate economic shifts.
*Average Growth Rate (avg_growth_rate):* Positive growth suggests a robust market; fluctuations may reflect changing conditions.
*Locality Count (locality_count):* An increasing count suggests market expansion or better data coverage.
The market_trends_by_year function provides a high-level view of Melbourne’s property market, aiding investors, analysts, and policymakers in understanding long-term trends.


## <u> Conclusion </u>

This vignette demonstrates how to use the package’s functions to analyze Melbourne’s property market. Whether examining year-over-year price changes, identifying top-performing suburbs, or evaluating market stability, these tools offer valuable insights for investors, analysts, and other stakeholders seeking a deeper understanding of Melbourne’s evolving property landscape.





